---
title: "hotpatchR Overview"
author: "David Munoz Tord"
date: "2026-04-15"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hotpatchR Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(hotpatchR)
```

# Introduction

This vignette explains the `hotpatchR` design and why runtime namespace patching is the right tool for legacy hotfix workflows.

## The legacy package lockdown problem

A loaded R package lives in a locked namespace. That means internal functions are resolved from inside the package bubble.
When one internal function calls another, R does not look in the global environment.

So if you fix `a_freq_j()` in the global environment, `tt_to_tlgrtf()` inside the package still calls the original broken version.

This is the namespace trap that makes legacy hotfixing so difficult.

## Why the global workaround fails

The usual workaround is:

- identify the broken function
- find all parent callers in the package
- copy the broken function and every dependent caller into a hotfix script
- source the script into the global environment
- run tests to confirm the patched path

This is painful because a bug in one internal function can require copying many internal callers, even when only one implementation actually needs to change.

## hotpatchR philosophy

Instead of pulling functions out into the global environment, `hotpatchR` performs surgical edits inside the package namespace.
That means:

- fix only the broken function
- internal callers automatically resolve to the updated implementation
- the package remains loaded and otherwise unchanged

## Basic workflow

```{r example}
# Simulate a locked package environment.
pkg_env <- new.env()
pkg_env$broken_child <- function() "I am broken"
pkg_env$parent_caller <- function() pkg_env$broken_child()
lockEnvironment(pkg_env)
lockBinding("broken_child", pkg_env)
lockBinding("parent_caller", pkg_env)

# Apply the patch.
fixed_child <- function() "I am FIXED"
inject_patch(pkg_env, list(broken_child = fixed_child))

pkg_env$parent_caller()
```

If `pkg_env` were a real package namespace, the same internal caller would now use the patched `broken_child()`.

## How inject_patch works

`inject_patch()`:

1. identifies the target namespace or environment
2. unlocks the binding for the named object
3. assigns the replacement function into that environment
4. re-locks the binding

Because the replacement function's parent environment is set to the target namespace, it can still call other internal objects from the same package.

## Rolling back a patch

If you need to restore the original binding, `undo_patch()` reverses the previous change.

```{r undo-example}
undo_patch(pkg_env, "broken_child")
pkg_env$parent_caller()
```

## Hotfix scripts

`apply_hotfix_file()` is a convenience wrapper for scripting hotfix application. A compatible hotfix script should define:

- `pkg` (optional, if not passed explicitly)
- `patch_list`, a named list of replacement functions

Example hotfix file:

```r
pkg <- "junco"
patch_list <- list(
  a_freq_j = function(...) {
    # fixed implementation
  }
)
```

Then apply it with:

```r
apply_hotfix_file("dev/junco_hotfix_v0-1-1.R")
```

## Next steps

The current package is focused on the core runtime patching path.
Future enhancements may include patch comparison, dependency scanning, and CI-friendly test wrappers that explicitly preserve the patched namespace during test execution.
